---
title: "Portfolio Optimization under Power Law Returns"
author: "Christopher McDonald"
date: "26 Aug 2022"
abstract: >
    Monte-Carlo simulation is used to evaluate the distribution
    of returns for portfolios of investments made at the seed stage.
    The results confirm that the main benefit of
    diversification is a higher typical (median) return multiplier.
    Under the simple model presented, gains from diversification
    reach diminishing returns at around 1000 firms for "early-exit"
    strategies, but continue to be substantial beyond this
    for "late-exit" strategies.
    Limitations of the model are briefly discussed.
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: "readable"
bibliography: references.bib
toc-title: "Table of Contents"
---

```{r notebook_settings, include=FALSE}
# set output device to svg
knitr::opts_chunk$set(dev = "svg")
knitr::opts_chunk$set(fig.topcaption = TRUE)
colourise <- function(x, color) {
    if (knitr::is_latex_output()) {
        sprintf("\\textcolor{%s}{%s}", color, x)
    } else if (knitr::is_html_output()) {
        sprintf("<span style='color: %s;'>%s</span>", color, x)
    } else {
        x
    }
}
```

```{css set_styles, echo=FALSE}
caption, .caption {
    padding-top: 8px;
    padding-bottom: 8px;
    color: #656565;
    text-align: left;
}
```

```{r setup_environment, include=FALSE}
# data manipulation and visualisation tools
library(tidyverse)

# tables
library(knitr)

# simulate from arbitrary models, represented as directed graphs
source("R/simulate.R")

# functions which generate data draw from various return distributions (used as
# inputs to simulate.R)
source("R/model.R")

# set seed for reproducibility
set.seed(123)
```

# Introduction

## Previous work

What is known about the distribution of returns in early-stage venture
investments?

The most comprehensive report I could find on this topic is the AngelList
report
_[Startup Growth and Venture Returns (Othman, 2019)](https://angel.co/pdf/growth.pdf)_.
I will summarise its methods, main results and the
conclusions of the author below.

The report uses AngelList's data on early-stage venture investments to study
the distribution of returns for "winning" investments - that is, investments
with a return multiplier >1.
They study how this distribution varies with 2 factors: (1) the duration of the
investment^[More precisely, the effective duration of the investment. See the
report for details.] and (2) the time that the initial investment is made
(e.g. seed stage, series A, series B etc).

The main findings of the report are:

* A power law (i.e. [Pareto](https://en.wikipedia.org/wiki/Pareto_distribution))
distribution appears to describe the distribution of multipliers well. In
particular, it fits the data better than a lognormal distribution.

* The thickness of the power law tail (determined by its parameter $\alpha$)
increases with the duration of the investment, and is higher for earlier
investments than later ones.

* In particular, the best fit model for seed stage investments with durations of
5 or more years is a Pareto distribution with a power $\alpha < 1$^[The report
defines a power law slightly differently to the standard definition of the
Pareto distribution. Here I will use the standard definition.]. This model
predicts an unbounded (i.e. infinite) expected multiplier.

* Startups tend to grow faster (in the sense of having a higher IRR) in their
earlier years, and slower in their later years. They provide empirical data to
back up this claim.

Based on these results, they make several conclusions:

* There is a qualitative difference between seed stage investments and later
investments; seed stage investments have theoretically infinite
multipliers, while the expected multipliers for later stage investments are
bounded.

* Seed stage investors can increase their expected multiplier via
diversification (that is, by adding firms to their portfolio).

* Seed stage investors "should invest in every credible deal", while later-stage
investors "should be more thoughtful and discerning" about their investments.

* The fundamental nature of startup stock (as an asset class) changes over its
lifetime.

## Outline

In [Theory], I argue that the infinite expected multiplier emphasized by
@othman_report is misleading.
Then in [Simulation Study] I explain the assumptions, methods and results
of the simulation study I performed.
Finally in [Discussion] I briefly discuss my main findings, and list
the weaknesses of my model.

# Theory

In absence of anything else, physics sets a hard upper limit on the maximum
achievable return for any venture^[It appears that the maximum value we can
produce in any given volume is finite. Since our speed of expansion is limited
by the speed of light, this implies rates of growth must be finite.].
This tells us that the infinite expected multipliers for seed-stage investments
predicted by @othman_report's model is misleading, and should not be taken literally.
(Common sense suggests that the real upper limit on multipliers are
much lower than the limit set by physics.)

This contradicts @othman_report's claim that seed-stage investments are
"qualitatively" (as well as "quantitatively") different to later-stage
investments.
Specifically, once you introduce a maximum feasible rate of return,
the overall expected multiplier must be finite.
However, it is likely that seed-stage investments are still
"quantitatively" very different to later-stage investments - for example,
they likely offer much higher expected multipliers.

To take into account the existence of a maximum possible return multiplier,
@othman_report's model (of Pareto-distributed multipliers) can be modified by
introducing a "cut-off" scale $x_{\text{max}}$.
The largest multiplier observed in Othman was 115x, while Peter Thiel
is said^[https://www.angellist.com/blog/what-angellist-data-says-about-power-law-returns-in-venture-capital]
to have achieved a multipler of 2200x on his investment in
Facebook [@othman_blog].
To be conservative, I will use $x_{\text{max}} = 10^6$,
which corresponds to a $10 billion return on an initial investment
of $10,000.

Power law relationships are notoriously hard to fit.
Some data may match a power law over a given range, but have tails
which drop off faster outside this range.
It is hard to distinguish whether this is the case, because
(as in this case) you may not have any observations from the
range you want to study.

# Simulation Study

@othman_report provides tentative data on the distribution of returns for individual
firms, as a function of the time of investment (e.g. seed stage, series A, ...)
and the duration of the investment.
Here I will use this data along with Monte-Carlo simulation methods to
examine the distribution of returns for various portfolios, conditional on
certain assumptions about how the portfolios were constructed.
This allows me to compare the (theoretical) returns and risk profiles of
different portfolios.

These simulations aim to provide a better understanding of what heavy-tailed
distributions mean for portfolios of investments, and what an optimal portfolio
might look like in this case.
While the model presented here is highly simplified, with more realistic
assumptions, such simulations could also be used to directly inform
decision-making.

In this section, I first discuss the assumptions the model relies on.
Following this, I explain briefly how the model was simulated, and then
present the results.

## Assumptions

The following simplifying assumptions were used:

* The number of firms that are available to invest in is effectively infinite.

* The distribution of multipliers for each firm is independent.

* An equal amount is invested in each firm.

* There is no marginal cost to investing in an extra firm.

* Each investment is made at the seed stage.

* A given firm succeeds (achieves a positive multiplier) with probability
$p = 0.5$. In other words, a firm's investment multiple $m$ has 
$\mathbb{P}(m \gt 1) = p = \frac{1}{2}$ and
$\mathbb{P}(m \le 1) = 1 - p  = \frac{1}{2}$.

* The multipliers of successful firms are Pareto-distributed
(or truncated Pareto -distributed) with scale parameter
$x_{\text{min}} = 1$, and a shape parameter $\alpha$ which
depends on the duration of the investment $t$ as shown below.

```{r display_power_law, echo = FALSE, results = "asis"}
power_law_parameter <- tibble(
    duration = 1:8,
    shape = c(5, 3, 2.3, 2.1, 2, 1.93, 1.86, 1.8) - 1,
)
kable(
    power_law_parameter,
    col.names = c("Investment Duration (years)", "Shape Parameter $\\alpha$"),
    caption = str_c(
        "Table 1: Shape parameter $\\alpha$ of the distribution of the",
        " multiplier $m$, as a function of investment duration.",
        " Source: Otham (2019), Figure 7."
    ),
    escape = FALSE
)
```

```{r plot_success_distribution, echo=FALSE, warning=FALSE, fig.cap="Figure 1: Distribution of multipliers for successful (m > 1)  firms as a function of investment duration."}
#' Density function for of a Pareto-distributed random variable
dpareto <- function(x, shape, scale) {
    shape * (x / scale)^(-shape) / x
}
power_law_parameter %>%
    expand_grid(x = seq(1, 5, length.out = 1e3)) %>%
    mutate(density = dpareto(x = x, shape = shape, scale = 1)) %>%
    ggplot(aes(x = x, y = density, colour = factor(duration))) +
    geom_line() +
    xlim(1, 5) +
    labs(
        x = "Multiplier",
        y = "Probability Density",
        colour = "Investment Duration\n(Years)"
    )
```

* The multiplier of failing firms is $\text{Beta}(a, b)$-distributed with
parameters $a = 3$ and $b = 1$ (corresponding to a mean of 0.75 and standard
deviation of 0.2). This distribution is shown below.

```{r plot_failure_distribution, echo=FALSE, fig.cap="Figure 2: Distribution of multipliers for failing (m < 1) firms. Mean multiplier is shown in red."}
shape1 <- 3
shape2 <- 1
ggplot() +
    geom_vline(xintercept = 0.75, col = "red", linetype = "dashed") +
    geom_function(fun = dbeta, args = list(shape1 = shape1, shape2 = shape2)) +
    labs(
        x = "Multiplier",
        y = "Probability Density"
    ) +
    lims(x = c(0, 1), y = c(0, 3))
```

* The distribution of returns is assumed to vary along 3 dimensions:
    1. The number of firms in the portfolio $k$: either a "small" portfolio
    ($k = 10$), a "medium" portfolio ($k = 100$) or a "large" portfolio
    ($k = 1000$).
    2. The duration of the investment $t$: "early exit" ($t = 2$ years) or
    "late exit" ($t = 8$ years).
    3. The base distribution $D$: the distribution that multipliers
    $m$ are drawn from. Either a standard Pareto distribution
    (the original model proposed by @othman_report), or a
    truncated Pareto distribution - i.e. a Pareto distribution whose tail
    is cut off at the maximum scale $x_{\text{max}} = 10^6$.

## Methods

The distribution of the multiplier $m$ for a given set of parameters was
estimated using Monte-Carlo simulation.
Specifically, the Monte-Carlo distribution for a given set of parameters
($k$, $t$, $D$) was calculated as follows:

* Randomly sample multipliers for each of the $k$ firms in the portfolio, based
on the base distribution $D$ and the investment duration $t$.
* Compute the overall multiplier of the $k$-firm portfolio by taking the mean of
the multiplier over the firms in the portfolio.
* Repeat this process $n$ times to gain $n$ samples from the distribution of 
$k$-firm portfolio multipliers. Here a relatively large sample size was chosen
($n = 10,000$) in order to ensure the tails are adequately sampled.

```{r setup_model, echo=FALSE}
model_graph <- graph_from_literal(success -+ multiplier) # styler: off, nolint
pareto_model <- list(
    success = success_simulator,
    multiplier = multiplier_simulator
)
truncated_pareto_model <- list(
    success = success_simulator,
    multiplier = truncated_multiplier_simulator
)
```

```{r setup_parameters, echo=FALSE}
# Number of Monte Carlo samples
n_samples <- 1e4

#' Generate sets of parameters
#'
#' Arguments are any parameter which may vary between models
collect_parameters <- function(exit_year) {
    list(
        success = list(
            success_probability = 0.5
        ),
        multiplier = list(
            # The power law parameter corresponding to exiting n years after an
            # initial investment at the seed stage, for n = 1, 2, ..., 8.
            power_law_parameter = c(5, 3, 2.3, 2.1, 2, 1.9, 1.8, 1.7) - 1,
            exit_year = exit_year,
            # Scale: maximum value of return. Only used in truncated model.
            scale = 1e6,
            shape1 = 3,
            shape2 = 1
        )
    )
}

# Full set of models to simulate
model_set <- expand_grid(
    k = 10^(1:3),
    exit_year = c(2, 8),
    model_class = list(
        `Pareto` = pareto_model,
        `Truncated Pareto` = truncated_pareto_model
    ),
) %>%
    mutate(model_parameters = map(exit_year, collect_parameters)) %>%
    mutate(
        model_function =
            map2(
                model_class,
                model_parameters,
                ~ partial(
                    sim_dag,
                    model_graph = model_graph,
                    simulators = .x,
                    parameters = .y
                )
            )
    ) %>%
    mutate(model_class = as_factor(names(model_class)))
```

## Results

```{r run_simulation, echo=FALSE}
# run Monte-Carlo simulation
model_results <- model_set %>%
    mutate(
        samples = map2(k, model_function, portfolio_simulator, n = n_samples)
    ) %>%
    unnest(samples) %>%
    select(!c(group_id, success_fraction))
```

A summary of the results is shown below.

```{r summarise_results, echo=FALSE}
results_summary <- model_results %>%
    group_by(k, exit_year, model_class) %>%
    summarise(
        Mean = mean(net_multiplier),
        Median = median(net_multiplier),
        `Standard Deviation` = sd(net_multiplier),
        `Mean Absolute Deviation` = mad(net_multiplier, constant = 1),
        .groups = "drop"
    ) %>%
    arrange(exit_year, k) %>%
    rename(
        `Investment Duration (Years)` = exit_year,
        `Portfolio Size` = k,
        `Model Type` = model_class
    )
kable(
    results_summary,
    digits = 3,
    escape = FALSE,
    caption = "Table 2: Summary of results."
)
```

In this case, the standard deviation mainly measures the thickness
of the tail, since it is heavily skewed by outliers.
As a result it is not especially useful as a measure of the typical
variability you expect when sampling from the distribution.
Instead, to measure variability we will use the mean absolute deviation (MAD),
which tells us how much a given sample will typically vary from the median.

Monte-Carlo distributions for the multipliers are plotted below.

```{r plot_early_exits, echo=FALSE, warning=FALSE, fig.cap="Figure 3: Monte Carlo distribution of the return multiplier for early exits (t = 2 years)."}
model_results %>%
    filter(exit_year == 2) %>%
    ggplot(aes(x = net_multiplier, fill = factor(k))) +
    geom_density(alpha = 0.5) +
    xlim(0, 2.5) +
    facet_wrap(vars(model_class)) +
    labs(
        x = "Overall Multiplier",
        y = "Probability Density",
        fill = "Portfolio Size"
    )
```

```{r plot_late_exits, echo=FALSE, warning=FALSE, fig.cap="Figure 4: Monte Carlo distribution of the return multiplier for late exits (t = 8 years)."}
model_results %>%
    filter(exit_year == 8) %>%
    ggplot(aes(x = net_multiplier, fill = factor(k))) +
    geom_density(alpha = 0.5) +
    xlim(0, 50) +
    facet_wrap(vars(model_class)) +
    labs(
        x = "Overall Multiplier",
        y = "Probability Density",
        fill = "Portfolio Size"
    )
```
Some basic observations are:

* For an investment duration of 2 years - corresponding a power of
$\alpha = 2$ - the results for the Pareto and Truncated Pareto models are
practically identical. This is expected, as the models only differ in the tails,
which are unlikely to be sampled (given the "lightness" of the tail implied
by the value of $\alpha$).

* For this investment duration, the expected multiplier (1.37, 1.17 annualised)
is independent of the portfolio size (as expected), and the median multiplier
approaches the mean as the portfolio size increases - it is within 2% of the
mean multiplier for a portfolio size of $k = 1000$.

* For an investment duration of 8 years, corresponding to a power of
$\alpha = 0.8$, the results for the Pareto and Truncated Pareto appear
similar, but differ substantially in the tails.
As expected, the Truncated Pareto model has a bounded mean (around 70x, or 1.7x
annualised), while under the Pareto model the expected multiplier
increases dramatically with portfolio size.
In contrast, the median multiplier is almost identical between the two
models (as expected if they differ only in the tails).

* While the variability of the multiplier (as measured by the MAD) decreases
with portfolio size in the "early-exit" case, variability _increases_ with
portfolio size in the "late-exit" case. This can be seen clearly in the figures
above. This is because (1) as the portfolio size increases, the chance of seeing
outliers from the tail increases; and (2) outliers are extreme enough to
substantially skew the average.

# Discussion

## Contributions

The main contributions of this report are as follows:

* It empirically confirms that the main benefit of diversification
(i.e. a larger portfolio) is in a higher typical (median)
multiplier.

* The model suggests that for "early-exit" strategies, a diversified portfolio
gives reduced variability, while for "late-exit" strategies, a diversified
portfolio gives _increased_ variability (at least up to portfolio sizes of
1000 firms).

* The model also suggests that "late-exit" strategies perform better
(annualised mean multipliers of 1.7) than "early-exit" strategies
(annualised mean multiplier of 1.17) on average.

* The results give some indication of how much diversification is optimal.
Under the simple model above, the gains from diversification begin to
reach diminishing returns at around $k = 1000$ firms in the early-exit case.
However in the late-exit case, the gains from diversification
continue to be substantial even as the size of the portfolio increases into
the thousands of firms.

* Contrary to @othman_report, I have argued (in [Theory]) that there is likely
an upper limit on possible returns, which results in a bounded expected
multiplier for any given investment.

* Given finite expected multipliers for investments in individual firms,
the expected multiplier of a portfolio is independent of its size - that is,
adding firms to a portfolio does not improve your expected multiplier
(contrary to what @othman_report says).
So I argue that, in theory, the only reason to invest in a portfolio
(of more than one firm) at all is to improve your risk profile.

* However, my results provide support for most of @othman_report's
practical recommendations.
For example, the large gains found from diversification when investing at the
seed stage suggest that "invest in every credible deal" is good advice for
this stage.

## Limitations

There are many weaknesses and limitations of the simulation scheme presented
here.
Future work could more accurately model reality by taking the following
factors into account:

* Investors have a limited pool of firms they are able to invest in.
* Firms vary, and investors have different information and beliefs about
different firms and their likely returns. (In other words, my model does not
take into account an investor's "edge".)
* There are costs (due to research, due diligence etc) to investing in a given
firm.
* Usually there will be multiple investment and exit events for a given firm,
rather than just one of each.
    * Investors may make follow-on investments.
    * Investors may adapt their investment strategy as they learn more about the
firms they are invested in, and their valuations change.
* Investors will likely allocate capital unevenly among firms.
* In reality, the returns of firms will be somewhat correlated rather
than independent.
* The seed stage is not the only stage of investment; further work should
look at investments at other stages.
* It would be useful to see how the distribution of the return multiplier
changes over the time.
* The data source for the model parameters (specifically the power law
power $\alpha$) needs to be checked for accuracy. For example, no confidence
intervals (or other measures of uncertainty) were given in the original report.

# References

<div id = "refs"></div>
